---
title: "Coding Assessment (Q11–20)"
author: "Tianze (Vincent) Luo"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: yes
    toc_float:
      collapsed: no
      smooth_scoll: no
    df_print: paged
    number_sections: no
    toc_depth: 4
    code_folding: show
    theme: cosmo
    #css: "style.css"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

# 0. Preparation

**Load Package**
```{r message=FALSE}
rm(list=ls())
library(tidyverse)
library(data.table)
library(GWASTools)

```

**Load Data**
```{r}
# two allele freq files
maf_case = fread("hw00-coding-assessment/maf_CASE.frq", 
                 header = TRUE, stringsAsFactors = FALSE, check.names = FALSE)

maf_control = fread("hw00-coding-assessment/maf_CONTROL.frq",
                    header = TRUE, stringsAsFactors = FALSE, check.names = FALSE)

# logistic reg output
gwas_logistic = fread("hw00-coding-assessment/pheno_results.assoc.logistic",
                      header = TRUE, stringsAsFactors = FALSE, check.names = FALSE)
```

# 1. Visualizaitons

## 1.1. Histogram of the **case** allele frequencies
```{r}
ggplot(aes(x=MAF), data = maf_case) + 
  geom_histogram(binwidth = 0.02, color="black", fill="grey") + 
  labs(title="Histogram of the case allele frequencies") +
  theme_bw()
```


## 1.2. Histogram of the **control** allele frequencies
```{r}
ggplot(aes(x=MAF), data = maf_control) + 
  geom_histogram(binwidth = 0.02, color="black", fill="grey") + 
  labs(title="Histogram of the control allele frequencies") +
  theme_bw()
```


## 1.3. Scatterplot of case-individuals’ allele frequencies (y axis) v.s. control-individuals’ allele frequencies (x axis)

```{r}
df_maf = data.frame(case = maf_case$MAF, 
                    control = maf_control$MAF)

ggplot(data=df_maf, aes(x=control, y=case)) + 
  geom_point(color="darkblue", alpha=0.5, size=2) + 
  labs(title = "Scatterplot of case-individuals’ allele frequencies (y axis) v.s.\n control-individuals’ allele frequencies (x axis)") + 
  theme_bw()

```


## 1.4. Scatterplot of −log₁₀(p values) (y axis) vs chr4 basepair position (x axis)

```{r}
gwas_logistic = gwas_logistic %>% mutate(neg_log10p = -log10(P))
head(gwas_logistic)

ggplot(data = gwas_logistic, aes(x= BP, y= neg_log10p)) + 
  geom_point(color="darkblue", size=1.2) + 
  
  labs(title="chr4 Manhattan Plot",
       x = "chr4", y = latex2exp::TeX("$-log_{10}$($p$-value)") ) + 
  theme_bw() + 
  theme(axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12))

```


# 2. More Reports

>18. Reports the marker and p value of the marker **with the lowest p value**; then

Answer: **rs35197316 (P = 3.016e-12)**

```{r}
# checks
str(gwas_logistic) #make sure of the data type
gwas_logistic$TEST %>% unique()

# sort the data frame by P
gwas_logistic %>% 
  arrange(P) %>%
  select(SNP, P)

```

>19. Loads the `snp_annot.RData` workspace and merges the p values into the SNP annotation data frame object `snp_annot_df` in such a way that
>20. The order of the SNPs in snp_annot_df is unchanged.

```{r}
# loads
load("/ix1/hugen2072-2025s/ca/snp_annot.RData", verbose = TRUE)

class(snp_annot_df)
typeof(snp_annot_df)
snp_annot_df@data


# Clean: need rsID as the key for merging
df_rsid.p = gwas_logistic %>% select(SNP, P)
df_rsid.p


# Merge: The order of the SNPs in snp_annot_df is unchanged
# ***snp_annot_df@data < df_rsid.p

#check the dup & other things in key
length(snp_annot_df@data$rsID)
length(df_rsid.p$SNP)
all(df_rsid.p$SNP %in% snp_annot_df@data$rsID)
# duplicated(snp_annot_df@data$rsID) %>% sum()
# duplicated(df_rsid.p$SNP) %>% sum()
# is.na(snp_annot_df@data$rsID) %>% sum()
# is.na(df_rsid.p$SNP) %>% sum()

updated_data = 
  left_join(x = snp_annot_df@data,
            y = df_rsid.p,
            by = join_by(rsID == SNP))


#check if order changed
all(updated_data$rsID == snp_annot_df@data$rsID) #TRUE


#### Update the data of the S4 object ####
pData(snp_annot_df) <- updated_data

str(snp_annot_df)
head(snp_annot_df@data)
```


```{r}
sessionInfo()
```
